Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex
reducing a bit the paper length

 In this section we compare the four different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). %\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM. We count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\end{itemize}

One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins brings modularity at the cost of some performance loss.

>>>>

 In this section we compare the four different implementations according to the following criteria:
% \begin{itemize}
% 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
%	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
%	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
%	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM. We count all the Slang implementations as well as the ones on Cog's RTL.
%	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
%	\item \emph{Optimization potential:}
%	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
%	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
%\end{itemize}

%Vspce to fit all on page.
\vspace{-0.1cm}
	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
\vspace{-0.1cm}
	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
\vspace{-0.1cm}
	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
\vspace{-0.1cm}
	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
\vspace{-0.1cm}
	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
\vspace{-0.1cm}
	\paragraph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
\vspace{-0.1cm}
	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\vspace{0.2cm}

Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used. VM recompilation slows slightly the development process, but the resulting code is usually faster to run. Likewise, the use of plugins brings modularity at the cost of some performance loss.
